---
title: "Variant Visualization"
author: "Jânice Roberta de Paula"
date: "2026-03-18"
output:
  pdf_document: default
  html_document: default
---

# NGS Variant Analysis Report

## 1. Objective
To perform processing and analysis of sequencing data in order to identify possible nucleotide variants relative to the reference sequence AF086833, using a standard bioinformatics pipeline. \

## 2. Analysis Pipeline

The data were processed following the classical steps of NGS analysis. \

*Quality Control* \
**Tool used:** \
fastp \

*Functions:* \
Adapter removal \
Filtering of low-quality reads \
Trimming of read ends \

*Read Alignment* \
**Tool used:** \
BWA \

*Reference used:* \
AF086833 sequence \

*Result:* \
Generation of a SAM file, which was later converted to BAM format. \

*Alignment Processing* \
**Tool used:**
SAMtools \

**Procedures:** \
SAM -> BAM conversion \
BAM sorting (sort) \
indexing (index) \
alignment statistics (flagstat) \

**Final alignment file:** \
final_sorted.bam \

# 3. Variant Detection
Two different algorithms were used to increase the robustness of the analysis. \
**Method 1** \
*Tool:* \
BCFtools \
*Output:* \
variants1.vcf \
*Number of variants detected:* \
0 variants \
**This result indicates that the algorithm did not find sufficient statistical evidence to call variants.**

**Method 2** \
*Tool:* \
FreeBayes \
*Output:* \
variants2.vcf \
*Number of variants detected:* \
2 variants \
**Differences between algorithms are expected due to distinct statistical models.**

# 4. Variant Visualization
The variants were visualized in Integrative Genomics Viewer (IGV) for manual validation. \
**Files loaded:** \
AF086833.fasta (reference) \
final_sorted.bam (alignment) \
variants2_filtered.vcf (filtered variants) \

# 5. Visualization Observations
*Visual inspection shows:* \
Two variant positions along the sequence (~6 kb of the reference) \
Presence of reads supporting alternative bases \
Moderate coverage in the detected regions \
The variants appear as colored vertical markers, indicating possible SNPs (Single Nucleotide Polymorphisms). \

# 6. Sequencing Coverage
**The Coverage track shows:** \
Irregular distribution of reads along the sequence \
Presence of regions with low coverage \
Alignment clusters in some specific regions \

**This pattern may indicate:** \
Targeted amplification \
Specific fragmentation or preferential alignment regions. \

# 7. Interpretation
**The results indicate that:** \
The dataset has low variant diversity \
Only two possible mutations were identified \
The results depend on the algorithm used \

**The fact that only FreeBayes detected variants may indicate:**
Low coverage \
Limited read support \
A conservative threshold used by BCFtools. \

# 8. Conclusion
**The analysis demonstrated that:** \
The NGS processing pipeline was executed correctly \
Read alignment was successfully performed \
Two potential variants were identified in the analyzed sequence \
Visual validation in IGV confirms the presence of nucleotide variation signals. \

**These results illustrate the importance of:** \
Using multiple variant-calling algorithms \
Visually validating detected variants. \

# 9. Dataset Description
The sequencing data analyzed in this study correspond to the dataset SRR1972877, obtained from the NCBI Sequence Read Archive (SRA). \

**Characteristics of the dataset:** \
**Sequencing platform:** Illumina \
**Data type:** Paired-end reads \
**File format:** FASTQ \
**Total number of reads:** 2707417

# 10. Quality metrics
**Total reads before filtering:**	5.414834 M \
**Q20 score before filtering:** 307.291682 M (56.188092%) \
**Q30 score before filtering:** 264.796929 M (48.417953%) \

**Total reads after filtering:** 1.295060 M \
**Q20 score after filtering:** 110.178145 M (89.026850%) \
**Q30 score after filtering:** 93.783607 M (75.779631%) \

**GC content:** 57,4% \

# 11. Alignment Statistics
**Total reads:**	1295061 \
**Mapped reads:**	23 \
**Mapping rate:** 0,00%	\

# 12. Variants detected
```{r, include = FALSE, message = FALSE, warning = FALSE}
# Criando a tabela
variants_detected <- data.frame(
  Position = c(5876, 5954),
  Reference = c("C", "T"),
  Alternative = c("T", "C"),
  Depth = c(2, 2),
  Caller = c("FreeBayes", "FreeBayes")
)
```

```{r, echo= FALSE, message = FALSE, warning = FALSE}
knitr::kable(variants_detected, caption = "Variants detected by FreeBayes")
```

# 13. Data Visualization
**Figures generated during the analysis:** \
Genome coverage plot \
Variant distribution (Manhattan plot) \
IGV genome browser visualization \

# 14. Limitations
**The analysis presents some limitations:** \
Relatively low sequencing coverage \
Small number of detected variants \
Dependence on the reference sequence \
Differences between variant-calling algorithms \

# 15. Reproducibility
**The analysis pipeline can be reproduced using the following tools:** \
FastQC \
Fastp \
BWA \
SAMtools \
BCFtools \
FreeBayes \
IGV \
R \
*Scripts used in the analysis are available in the project directory.*

# 16. Computational Environment
The analysis was performed in a Linux environment. \
**Tools and versions used:** \
```{r, include = FALSE}
version_tools <- data.frame(
        Tools = c("Fastp", "FastQC", "BWA", "SAMtools", "Freebayes", "BCFtools", "VCF"),
        Version = c("1.1.0", "0.12.1", "0.7.17", "1.23", "1.3.6", "1.21", "4.3")
)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
knitr::kable(version_tools, caption = "Version Tools")
```


